---
title: "Dataset Builder"
author: "Greg Macfarlane"
date: "1/20/2020"
output: html_document
---

```{r setup, include=FALSE, cache = FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
library(sf)
library(leaflet)
library(tidyverse)
library(tigris)
library(tidycensus)
options(readr.show_progress = FALSE)
```

Constructing a PopulationSim implementation is not a trivial task. In this
document, we guide the users through constructing a basic PopulationSim
scenario. 

PopulationSim allows users to create controls at multiple levels. In this case
we are going to create three levels of controls:

  1. Region-wide: total population
  2. Tract-level: household size and income, and person age
  3. TAZ level: number of households
  
We are going to build this synthetic population for Box Elder, Weber, Davis, Salt Lake, and Utah Counties in
Utah. This involves the ranges of PUMAs shown below. We can get the census tracts these PUMAs
contain from the Census Bureau's Tract / PUMA lookup file. 

```{r puma}
puma_url <- "https://www2.census.gov/geo/docs/maps-data/data/rel/2010_Census_Tract_to_2010_PUMA.txt"
puma_tract <- read_csv(puma_url, col_types = list(PUMA5CE = col_integer())) %>%
  filter(STATEFP %in% c(49)) %>% 
  transmute(
    TRACT = str_c(STATEFP, COUNTYFP, TRACTCE),  
    PUMA = PUMA5CE
  ) %>%
  filter(PUMA %in% c(49001:49004, 11001:11002, 35001:35009, 57001:57002, 03001:03004))
puma_tract$PUMA <- formatC(puma_tract$PUMA, width = 5, flag = "0")
  
```

We also want to get the geographic boundaries layer for these tracts. We get
this directly from the Census.

```{r tracts, message = FALSE}
tr <- tracts("UT", county = unique(substr(puma_tract$TRACT, 3, 5)), 
             class = "sf", progress_bar = FALSE) %>%
  st_transform(4326) %>%
  transmute(GEOID) %>%
  left_join(puma_tract, by = c("GEOID" = "TRACT")) %>%
  filter(!is.na(PUMA))
```

The last step we need is the list of TAZs that intersect with these tracts. We
use a TAZ file obtained from the Mountainland Association of Government travel 
model. We filter the tracts based on whether they intersect.

```{r taz}
taz <- st_read("inputs/taz.geojson") %>% 
  mutate(TAZ = as.character(TAZID)) %>%
  # remove external stations
  filter(CO_FIPS > 0) %>%
  select(TAZ) %>%
  
  # remove taz that do not map to a tract
  st_join(tr) %>% filter(!is.na(PUMA)) %>%
  
  # remove duplicates of TAZ that map to many tracts
  group_by(TAZ) %>% slice(1) %>% ungroup()
```

The map below shows the TAZ and Tract boundaries for our controls. 
As the map shows, there is not a perfect correlation between the boundaries of the
TAZ and tract layers. Nevertheless to make the geographic crosswalk we will
use the simplifying assumption that each TAZ belongs to the tract that contains
the TAZ centroid. The TAZ covering Utah Lake will be removed when we do this.

```{r tazmap}
leaflet() %>%
  addProviderTiles(providers$Esri.WorldGrayCanvas) %>%
  addPolygons(data = tr, group = "Tract", color = "blue") %>%
  addPolygons(data = taz, group = "TAZ", color = "red") %>%
  addLayersControl(overlayGroups = c("Tract", "TAZ"))
```

## Geographic Cross-walk

To make the crosswalk we locate the taz centroids within each tract, and then
append the PUMA, county, and region codes. The plot below visualizes these points
as well as the TAZ, tract, and PUMA ids. The resulting table is the geographic 
crosswalk required by PopulationSim.

```{r crosswalk}
crosswalk <- taz %>%
  # determine which tract the taz is in
  select(TAZ) %>% st_centroid() %>%
  st_join(tr, st_within) %>% filter(!is.na(GEOID)) %>%
  # fill out remainder of columns
  rename(TRACT = GEOID) %>% mutate(REGION = "1")

plot(crosswalk %>% mutate(TAZ = as.numeric(TAZ)))
```

```{r print_crosswalk}
(crosswalk <- crosswalk %>% st_set_geometry(NULL) %>% as_tibble())
```


## Controls

Attributes we want to control for include:
 - Number of households
 - Household size
 - Household income
 - Age of persons
 - Household workers
 - Person worker status

### TAZ-level controls

The only TAZ-level control we get from WFRC is the number of households in each
TAZ. Therefore, this will be the only control in this file. The MAG data is only
available for 2020, so that's what we will be using even if 2017 or 2018 might
be more appropriate. We filter the list of controls to the TAZ that we are
getting in the crosswalk. We need to ensure that the TAZ code is a character
string and the household target is an integer.

```{r taz_control}
# read socioeconomic data file
(taz_control <- read_csv("inputs/Household_Projections_(TAZ).csv") %>% 
  transmute(TAZ = as.character(TAZ2), HHBASE = as.integer(ceiling(YEAR2020))) %>% 
  filter(TAZ %in% crosswalk$TAZ))
```

### Tract-Level Controls
The attributes available from the ACS include the following:
  - Household size, derived from Table `B08202: HOUSEHOLD SIZE BY NUMBER OF WORKERS IN HOUSEHOLD`
  - Household workers, derived from the same table
  - Age, derived from Table `B01001: SEX BY AGE`
  - Income, derived from Table `B19001:HOUSEHOLD INCOME IN THE PAST 12 MONTHS (IN 2018 INFLATION-ADJUSTED DOLLARS)` 

First, we can collect a unique list of all the tracts we want to get controls for,
alongside a complete list of all ACS variables
```{r tract_list}
mytracts <- unique(crosswalk$TRACT)
mycounties <- unique(substr(crosswalk$TRACT, 3, 5))
acsvars <- load_variables(2018, "acs5", cache = TRUE)
```

#### Household size and workers
Because these two values come in the same table, we will build them together.
In both cases the top-line category contains all households with that many
workers / persons or more.

```{r sizework}
swvars <- str_c("B08202_", sprintf("%03d", c(2:5, 6, 9, 13, 18)))
raw_sw <- get_acs("tract", variables = swvars, state = "UT", county = mycounties)

size_work <- raw_sw %>% 
  left_join(acsvars, by = c("variable" = "name")) %>%
  separate(label, c("VAR", "total", "label"), sep = "!!") %>%
  select(GEOID, label, estimate) 

works <- size_work %>%
  filter(grepl("work", label)) %>%
  mutate(
    num_work = str_extract(label, "\\d+"),
    workcat = case_when(
      num_work == 1 ~ "HHWORK1",
      num_work == 2 ~ "HHWORK2",
      num_work == 3 ~ "HHWORK3",
      TRUE ~ "HHWORK0"
    )
  ) %>% 
  group_by(GEOID, workcat) %>% summarize(count = as.integer(sum(estimate)))

sizes <- size_work %>%
  filter(!grepl("work", label)) %>%
  mutate(
   num_size = str_extract(label, "\\d+"),
   sizecat = str_c("HHSIZE", num_size) 
  ) %>%
  group_by(GEOID, sizecat) %>% summarize(count = as.integer(sum(estimate)))
```


#### Age
This is the number of people in each age category.

```{r age}
agevars <- str_c("B01001_", sprintf("%03d", c(3:25, 27:49)))
raw_ages <- get_acs("tract", variables = agevars, state = "UT", county = mycounties)
  
ages <- raw_ages %>%
  left_join(acsvars, by = c("variable" = "name")) %>%
  separate(label, c("VAR", "total", "sex", "age"), sep = "!!") %>%
  select(GEOID, sex, age, estimate)  %>%
  
  # regroup age categories
  mutate(
    numage = as.numeric(substr(age, 1, 2)),
    agecat = case_when(
      numage %in% c(15:24) ~ "PAGE1",
      numage %in% c(25:54) ~ "PAGE2",
      numage %in% c(55:64) ~ "PAGE3",
      numage %in% c(65:99) ~ "PAGE4",
      TRUE ~ "PAGE0" # children less than 15 not categorized in demo
    )
  ) %>%
  
  # consolidate men and women
  group_by(GEOID, agecat) %>%
  summarise(count = as.integer(sum(estimate)))
```

#### Income
This is the household income variable, which is categorized as follows:
  - <$15k
  - \$15k - \$30k
  - \$30k - \$60k
  - > $60k

```{r income}
incvars <- str_c("B19001_", sprintf("%03d", c(2:17)))
raw_incs <- get_acs("tract", variables = incvars, state = "UT", county = mycounties)

incs <- raw_incs %>%
  left_join(acsvars, by = c("variable" = "name")) %>%
  separate(label, c("VAR", "total", "income"), sep = "!!") %>%
  select(GEOID, income, estimate)  %>%
  # regroup income categories
  mutate(
    numinc  = stringr::str_extract(income, "\\d+"),
    inccat = case_when(
      numinc <  15 ~ "HHINC1",
      numinc <  30 ~ "HHINC2",
      numinc <  60 ~ "HHINC3",
      numinc >= 60 ~ "HHINC4",
      TRUE ~ as.character(NA)
    )
  ) %>%
  group_by(GEOID, inccat) %>%
  summarise(count = as.integer(sum(estimate)))
```


#### Tract Controls File
When all of the controls have been gathered, we can put them into one large table.

```{r tract_controls}
(tract_controls <- tibble(TRACT = mytracts) %>%
   left_join(ages  %>% spread(agecat,  count), by = c("TRACT" = "GEOID")) %>%
   left_join(incs  %>% spread(inccat,  count), by = c("TRACT" = "GEOID")) %>%
   left_join(works %>% spread(workcat, count), by = c("TRACT" = "GEOID")) %>%
   left_join(sizes %>% spread(sizecat, count), by = c("TRACT" = "GEOID"))
)
```

### Meta Controls
PopulationSim requires (we think) at least some region-level controls. We will
simply sum up the total population in the controls data to work with this.

```{r meta}
(meta <- tract_controls %>%
  summarise(
    REGION = 1, 
    totalPOP = sum(PAGE0) +  sum(PAGE1) +  sum(PAGE2) + 
      sum(PAGE3) +  sum(PAGE4)))
```

## PUMS Seed Data

The uncompressed seed data is too large to commit, so we have committed
compressed versions of the original files. We need to clean and re-save the PUMS
seed data files for households and for people. In the case of households,
this entails filtering the households without people, and replacing `NA` values
in the controls with other values.

```{r pums_hh, message=FALSE}
# read in pums HH file, specifying datatypes for key columns
pums_hh <- read_csv(
  "inputs/psam_h49.csv.zip",  
  col_types = list(SERIALNO = col_character(), NP = col_integer(), 
                   FINCP = col_number(), ADJINC = col_number(),
                   WGTP = col_number())
)

#' A function to replace NA values with non-missing stupid numbers
replace_na <- function(x) {
  ifelse(is.na(x),-8,x)
}

seed_hh <- pums_hh %>%
  # remove empty households
  filter(NP > 0) %>% filter(WGTP > 0) %>%
  # remove households from outside the region
  filter(PUMA %in% crosswalk$PUMA) %>%
  mutate(
    # create unique hh_id field
    hh_id = 1:nrow(.),
    # compute adjusted numeric income
    HHINCADJ = FINCP * ADJINC
  ) %>%
  # apply replacement function to key variables
  mutate_at(.vars = vars(NP, WIF, WGTP, HHINCADJ), replace_na)
```

```{r pums_persons, message = FALSE}
pums_persons <- read_csv(
  "inputs/psam_p49.csv.zip", 
  col_types = list(SERIALNO = col_character(), PWGTP = col_number())
)

seed_per <- pums_persons %>%
  # join hhid field, and only keep households we filtered down to.
  inner_join(seed_hh %>% select(SERIALNO, hh_id, WGTP)) %>%
  # replace NA values with something less stupid
  mutate_at(.vars = vars(PWGTP, AGEP), replace_na)
```




## Write out to files

With all of the data compiled, we now write out the files to the `data/` folder

```{r write out}
# Controls
write_csv(meta, "data/control_totals_meta.csv")
write_csv(tract_controls, "data/control_totals_tract.csv")
write_csv(taz_control, "data/control_totals_taz.csv")

# Seed
write_csv(seed_hh, "data/seed_households.csv")
write_csv(seed_per, "data/seed_persons.csv")

# Crosswalk
write_csv(crosswalk, "data/geo_cross_walk.csv")
```


